
---
title: Ofav 2024  survival analysis
author: "AS Huffmyer"
date: '2024'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(readxl)
library(ggsurvfit)
library(gtsummary)
library(cardx)
library(cowplot)
```

# Format data from entry. 

```{r}
library("tidyverse"); library("janitor"); library("readxl")

path <- "data/survival/scoring_data.xlsx"
data <- path %>%
  excel_sheets() %>%
  set_names() %>% 
  map_df(~ read_excel(path = path, sheet = .x)%>%mutate(across(.fns = as.character)))
# can create df versions for 'resurrection' and 'lemon' analysis from "data" when/if desired

# remove rows (larvae) with specific ambiguous phenotype (lemons), allocation errors (miss), and pipet/handling process death (died_dried) to make clean ready to analyze survivorship dataset
good <-data %>% #filter (miss=="0",lemon=="0", died_dried=="0") %>% 
  select(-miss, -resurrection, -lemon, -died_dried, -two, -settle)%>%
  select(-row,-column) %>% #initial allocation, not actual assessment
  pivot_longer(cols=-c("plate","well", "conical", "ramp-treatment", "stress-treatment"),names_to ="time", values_to ="value")%>%
  mutate(plate=as.factor(plate)) 

#write_rds(good, "output/survival/survivorship")
write_csv(good, "output/survival/survivorship.csv")
```

# Read data 

Read in data. 
```{r}
data<-read_csv("output/survival/survivorship.csv")
```

Format. 
```{r}
data<-data%>%
  mutate(time=as.numeric(time))
```

Remove outlier plate 51
```{r}
data<-data%>%
  filter(!plate=="51")%>%
  filter(!plate=="65")%>%
  droplevels()
```

Create data frames with control and high temperature subsets. 
```{r}
ambient<-data%>%filter(`stress-treatment`=="ambient")
stress<-data%>%filter(`stress-treatment`=="stress")
```

Check lengths. 
```{r}
length(unique(ambient$plate))
length(unique(stress$plate))
```

Rename to status. 
```{r}
ambient<-ambient%>%
  rename(status=value, ramp_treatment=`ramp-treatment`, stress_treatment=`stress-treatment`)

stress<-stress%>%
  rename(status=value, ramp_treatment=`ramp-treatment`, stress_treatment=`stress-treatment`)

ambient<-as.data.frame(ambient)
stress<-as.data.frame(stress)

ambient$ramp_treatment<-as.factor(ambient$ramp_treatment)
ambient$stress_treatment<-as.factor(ambient$stress_treatment)
ambient$conical<-as.factor(ambient$conical)
ambient$plate<-as.factor(ambient$plate)
ambient$well<-as.factor(ambient$well)

stress$ramp_treatment<-as.factor(stress$ramp_treatment)
stress$stress_treatment<-as.factor(stress$stress_treatment)
stress$conical<-as.factor(stress$conical)
stress$plate<-as.factor(stress$plate)
stress$well<-as.factor(stress$well)
```


# Generate Kaplan Meier survival curves 

## Ambient stress

```{r}
s1 <- survfit(Surv(time, status) ~ ramp_treatment, data = ambient)
str(s1)
```

Plot the survival function
```{r}
survfit2(Surv(time, status) ~ ramp_treatment, data = ambient, conf.type="plain", conf.int=0.95) %>% 
  ggsurvfit() +
  labs(
    x = "Time point",
    y = "Survival probability"
  )
```

Use a log rank model to determine statistical differences in curves. 
```{r}
survdiff(Surv(time, status) ~ ramp_treatment, data = ambient)

# Call:
# survdiff(formula = Surv(time, status) ~ ramp_treatment, data = ambient)
# 
#                           N Observed Expected (O-E)^2/E (O-E)^2/V
# ramp_treatment=ambient 1440       29     54.5      11.9      25.3
# ramp_treatment=ramp    1440       80     54.5      11.9      25.3
# 
#  Chisq= 25.3  on 1 degrees of freedom, p= 5e-07 
```
Significant difference in survival. 

Analyze again with a Cox proportional hazards model. 
```{r}
coxph(Surv(time, status) ~ ramp_treatment, data = ambient)

# Call:
# coxph(formula = Surv(time, status) ~ ramp_treatment, data = ambient)
# 
#                      coef exp(coef) se(coef)     z       p
# ramp_treatmentramp 1.0429    2.8376   0.2168 4.811 1.5e-06
# 
# Likelihood ratio test=26.28  on 1 df, p=2.958e-07
# n= 2880, number of events= 109  

coxph(Surv(time, status) ~ ramp_treatment, data = ambient) %>% 
  tbl_regression(exp = TRUE) 
```

```{r}
library(survminer)
library(ggplot2)

ggsurvplot(s1, conf.int = 0.95, censor= F,
           ggtheme = theme_minimal())
```

# Generate Kaplan Meier survival curves 

## Stress temperature

```{r}
s2 <- survfit(Surv(time, status) ~ ramp_treatment, data = stress)
str(s2)
```

Plot the survival function
```{r}
survfit2(Surv(time, status) ~ ramp_treatment, data = stress) %>% 
  ggsurvfit() +
  labs(
    x = "Time Point",
    y = "Survival probability"
  )
```

Use a log rank model to determine statistical differences in curves. 
```{r}
survdiff(Surv(time, status) ~ ramp_treatment, data = stress)

# Call:
# survdiff(formula = Surv(time, status) ~ ramp_treatment, data = stress)
# 
#                           N Observed Expected (O-E)^2/E (O-E)^2/V
# ramp_treatment=ambient 1152       47     32.4      6.53      12.3
# ramp_treatment=ramp    1440       26     40.6      5.22      12.3
# 
#  Chisq= 12.3  on 1 degrees of freedom, p= 5e-04  
```

Analyze again with a Cox proportional hazards model. 
```{r}
coxph(Surv(time, status) ~ ramp_treatment, data = stress)

# Call:
# coxph(formula = Surv(time, status) ~ ramp_treatment, data = stress)
# 
#                       coef exp(coef) se(coef)      z        p
# ramp_treatmentramp -0.8340    0.4343   0.2444 -3.412 0.000645
# 
# Likelihood ratio test=12.27  on 1 df, p=0.0004598
# n= 2592, number of events= 73 

coxph(Surv(time, status) ~ ramp_treatment, data = stress) %>% 
  tbl_regression(exp = TRUE) 
```

```{r}
ggsurvplot(s2, conf.int = 0.95, censor= F,
           ggtheme = theme_minimal())
```


# Generate plots 
 
Set theme. 
```{r}
my_theme<-theme_classic()
```

ambient
```{r}
plot1<-survfit2(Surv(time, status) ~ ramp_treatment, data = ambient) %>% 
  ggsurvfit() +
  labs(
    x = "Time Point",
    y = "Survival probability",
    title="Ambient"
  )+
  scale_color_manual(values=c("blue", "red"))+
  scale_fill_manual(values=c("blue", "red"))+
  ylim(0.6,1)+
  my_theme+
  theme(legend.position="none");plot1

plot1<-plot1+add_confidence_interval();plot1
```

Ambient by plates 
```{r}
plot1b<-survfit2(Surv(time, status) ~ plate, data = ambient) %>% 
  ggsurvfit() +
  labs(
    x = "Time Point",
    y = "Survival probability",
    title="Ambient"
  )+
  #scale_color_manual(values=c("blue", "red"))+
  #scale_fill_manual(values=c("blue", "red"))+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot1b

plot1b+add_confidence_interval()
```

Stress
```{r}
plot2<-survfit2(Surv(time, status) ~ ramp_treatment, data = stress) %>% 
  ggsurvfit() +
  labs(
    x = "Time Point",
    y = "Survival probability",
    title = "Stress", 
    legend = "Ramp Treatment"
  )+
  ylim(0.6, 1)+
  scale_color_manual(values=c("blue", "red"))+
  scale_fill_manual(values=c("blue", "red"))+
  my_theme+
 # geom_text(x=10, y=0.2, label="Cox PH p=0.025")+
  #geom_text(x=10, y=0.15, label="Temperature vs control p=0.002")+
  #geom_text(x=10, y=0.10, label="All others vs control p>0.05")+
  theme(legend.position="right");plot2

plot2<-plot2+add_confidence_interval();plot2
```

Stress by plates. 
```{r}
plot2b<-survfit2(Surv(time, status) ~ plate, data = stress) %>% 
  ggsurvfit() +
  labs(
    x = "Time Point",
    y = "Survival probability",
    title = "Stress Treatment", 
    legend = "Ramp Treatment"
  )+
  #ylim(0.8, 1)+
  #scale_color_manual(values=c("blue", "red"))+
  #scale_fill_manual(values=c("blue", "red"))+
  my_theme+
 # geom_text(x=10, y=0.2, label="Cox PH p=0.025")+
  #geom_text(x=10, y=0.15, label="Temperature vs control p=0.002")+
  #geom_text(x=10, y=0.10, label="All others vs control p>0.05")+
  theme(legend.position="right");plot2b

plot2b+add_confidence_interval()
```

Assemble plot
```{r}
plots<-plot_grid(plot1, plot2, rel_widths=c(0.75,1), ncol=2)

ggsave(plots, filename="figures/KMcurves.png", width=10, height=4)
```


